\bf Calendar-based graphics for visualizing people's daily schedules

### 2019-10-15 reviews/response.Rmd

---
title: Response to reviewers
author: JCGS-18-293-R1
output:
  MonashEBSTemplates::memo:
    includes:
      in_header: preamble2.tex
fontfamily: mathpazo
---

We thank the reviewers for their helpful feedback which has improved the paper. The point by point description of changes are below: the reviewers' comments are in red and our response is in black.

@@

> Abstract: "modular arithmetic" is more accurate. Or omit the sentence altogether. I agree with comments  of reviewer 1 on this topic.

Thanks. We've revised to "modular arithmetic".

> Page 2 par 2 first sentence. There are other broader applications than this, eg, pollution monitoring. Last sentence "but conventional displays" (omit "the")

We've rephrased "people's daily schedules" to "human activities" that would include pollution monitoring.

@@

> Par 2. rephrase "more easily pop-up to viewers".

We've rephrased to "to accentuate unusual patterns, such as those related to special events, for viewers".

> Page 7. Second line. You might want to add that fifth row could be blank. Wrapping around the last few days could be confusing. Could this be made optional?

We have opened a github ticket on this issue, which will be implemented in the future.

@@

> Line 36, what are M and N? You explain them later, but should be at first usage.

Done.

> Page 8. I agree with reviewer 1. These figures are not needed. Move to supplementary material.

We dropped figure 4 and 5 (diagrams for laying out the calendar).

@@

> Figure 6. Could the figure have some indicator of the wrapping of few days to first row? It would help the reader.

We noted this in the figure caption. If adding some indicators, these indicators would possibly receive more attentions from the viewers, thus deemphasizing the main message.

@@

> In your figures is global scaling used, unless otherwise stated?

Yes.

> Page 17, figure 10. I don't see negative correlation. Just two slopes, both positive. Am I missing something?

We deleted the sentence, "This is due to the morning and afternoon commuter patterns in Figure 6: the adjacent hours are positively correlated when approaching to the peak hour but negatively correlated when moving away from the peak.", to mitigate the confusion.

@@

Done.

# Reviewer 1 {-}

> "Would it be possible to have a plot for each day in a calendar plot with thick lines separating the months?  That kind of plot is common for calendar heatmaps."  I think that is important.  Perhaps they can build it in.  If not, why not?

The white space between months already emphasises the separations between months. The lines separating the months are also thicker than the ones separating the days in the calendar plot.

@@

> p17 Figure 10  Why not align by days of the week?

TODO

> p19 Text implies that more than one date can be selected in the Shiny app.  The app was kludgy and did not work smoothly.  It is a nice idea, but interaction has to be fast and work well.  Zooming did not work properly and was not able to link from the weather plots.  Fix or do not mention in article.

We deleted the sentence "Conversely, selecting on weather data plots, linked to the calendar can help to assess if very hot/cold days and heavy rain, affect the number of people walking in downtown Melbourne." from Section 2.3.4.

### 2019-10-15 calendar-vis.Rmd
addressed the rest of the comments with Di

The purpose of the calendar-based visualization is to provide insights into people's daily schedules, relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia [@ped]. There are many sensors installed across the inner-city area, that count pedestrians every hour (Figure \@ref(fig:ped-map)). Data from 43 sensors in 2016 is analyzed here. This data can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. Patterns relative to special events (such as public holidays and recurring cultural/sporting events) would be worth studying in comparison to regular days, but conventional displays of time series data may bury this detail.

@@

The algorithm of transforming data for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month conforms to a layout of 5 rows and 7 columns, where rows and columns refer to the week of month and the day of week respectively. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is deterministic given the year of the calendar. Months are of different lengths, ranging from 28 to 31 days. Some months could extend over six weeks, but for these months the last few days are wrapped up to the top row of the block for compactness, and because it is convention. The notation for creating these cells is as follows:

@@

```{r month-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "(ref:month-diagram-cap)"}
include_graphics("img/month.png")
```

(ref:month-diagram-cap) Illustration of the indexing layout for cells in a month, where $k$ is day of the week, $g$ is day of the month, $(i, j)$ indicates grid position.

@@

```{r year-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "(ref:year-diagram-cap)"}
include_graphics("img/year-diagram.png")
```

(ref:year-diagram-cap) Illustration of the indexing layout for months of one year, where $M$ and $N$ indicate number of rows and columns, $b$ is a space parameter separating cells.

@@

(ref:fs-2016-cap) The calendar plot of hourly foot traffic at Flagstaff Station, using line graphs. The disparities between weekday and weekend along with public holiday, are immediately apparent. The arrangement of the data into a $3 \times 4$ monthly grid represents all the traffic in 2016. Note that the algorithm wraps the last few days in the sixth week to the top row of each month block for a compact layout, which occurs in May and October.

@@

Almost any type of plot can be shown in a calendar pane. Most of the full range of plotting capabilities in **ggplot2** is available. An example is shown in Figure \@ref(fig:scatterplot): the panes contain lag scatterplots, constructed with local scaling for each day at Flagstaff Station, where the lagged hourly count is assigned to the `x` argument and the current hourly count to the `y` argument. It indicates strong autocorrelation on weekends, and weak autocorrelation on work days. The V-shape in the weekday graphs arises when the next hour sees a substantial increase or decrease in counts. This is due to the morning and afternoon commuter patterns in Figure \@ref(fig:fs-2016): the adjacent hours are positively correlated when approaching to the peak hour but negatively correlated when moving away from the peak.

```{r scatterplot, fig.height = 5, fig.cap = "(ref:scatterplot-cap)"}
```

(ref:scatterplot-cap) Lag scatterplots, with local scaling, in the daily calendar layout. Each hour's count is plotted against the previous hour's count at Flagstaff Station, to demonstrate the autocorrelation at lag 1. The correlation between them is more consistent on non-work days than work days.

@@

Linking calendar displays to other types of charts is valuable to visually explore the relationships between variables. An example can be found in the **wanderer4melb** shiny application [@R-wanderer4melb]. The calendar most naturally serves as a tool for date selection: by selecting and brushing the glyphs in the calendar, it subsequently highlights the elements of corresponding dates in other time-based plots. Conversely, selecting on weather data plots, linked to the calendar can help to assess if very hot/cold days and heavy rain, affect the number of people walking in downtown Melbourne. The linking between weather data and calendar displays is achieved using the common dates.

@@

\newpage

@@

(ref:hod-cap) Scatterplots of half-hourly energy usage by time of day, with hourly averages overlaid, faceted by household and type of day. All households have different weekdays versus weekends daily routines. On weekdays, household 1 wakes up early before 6, and household 2 around 6, followed by household 3 and 4. The use of air conditioning is notable in households 1 and 2, as seen by horizontal clusters of points.

>>>>

The purpose of the calendar-based visualization is to provide insights into human activities, especially relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia [@ped]. There are many sensors installed across the inner-city area, that count pedestrians every hour (Figure \@ref(fig:ped-map)). Data from 43 sensors in 2016 is analyzed here. This data can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. Patterns relative to special events (such as public holidays and recurring cultural/sporting events) would be worth studying in comparison to regular days, but conventional displays of time series data may bury this detail.

@@

The algorithm of transforming data for constructing a calendar plot uses modular arithmetic, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month conforms to a layout of 5 rows and 7 columns, where rows and columns refer to weeks of the month and days of the week respectively. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is deterministic given the year of the calendar. Months are of different lengths, ranging from 28 to 31 days. Some months could extend over six weeks, but for these months the last few days are wrapped up to the top row of the block for compactness, and because it is convention. The notation for creating these cells is as follows:

@@

<!--
```{r month-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "(ref:month-diagram-cap)"}
include_graphics("img/month.png")
```

(ref:month-diagram-cap) Illustration of the indexing layout for cells in a month, where $k$ is day of the week, $g$ is day of the month, $(i, j)$ indicates grid position.
-->

@@

<!--
```{r year-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "(ref:year-diagram-cap)"}
include_graphics("img/year-diagram.png")
```

(ref:year-diagram-cap) Illustration of the indexing layout for months of one year, where $M$ and $N$ indicate number of rows and columns, $b$ is a space parameter separating cells.
-->

@@

(ref:fs-2016-cap) The calendar plot of hourly foot traffic at Flagstaff Station ranging from 0 to 6952, using line graphs. The disparities between weekday and weekend along with public holiday, are immediately apparent. The arrangement of the data into a $3 \times 4$ monthly grid represents all the traffic in 2016. Note that the algorithm wraps the last few days in the sixth week to the top row of each month block for a compact layout, which occurs in May and October.

@@

Almost any type of plot can be shown in a calendar pane. Most of the full range of plotting capabilities in **ggplot2** is available. An example is shown in Figure \@ref(fig:scatterplot): the panes contain lag scatterplots, constructed with local scaling for each day at Flagstaff Station, where the lagged hourly count is assigned to the `x` argument and the current hourly count to the `y` argument. It indicates strong autocorrelation on weekends, and weak autocorrelation on work days. The V-shape in the weekday graphs arises when the next hour sees a substantial increase or decrease in counts.

```{r scatterplot, fig.height = 11, out.width = "75%", fig.cap = "(ref:scatterplot-cap)"}
```

(ref:scatterplot-cap) Lag scatterplots, with local scaling, in the daily calendar layout. Each hour's count is plotted against the previous hour's count at Flagstaff Station, to demonstrate the autocorrelation at lag 1. The correlation between them is more consistent on non-work days (blue) than work days (red).

@@

Linking calendar displays to other types of charts is valuable to visually explore the relationships between variables. An example can be found in the **wanderer4melb** shiny application [@R-wanderer4melb]. The calendar most naturally serves as a tool for date selection: by selecting and brushing the glyphs in the calendar, it subsequently highlights the elements of corresponding dates in other time-based plots. The linking between weather data and calendar displays is achieved using the common dates.

@@

@@

(ref:hod-cap) Scatterplots of half-hourly energy usage by time of day, with hourly averages overlaid, faceted by household and type of day. On weekdays, household 1 wakes up early before 6, and household 2 around 6, followed by household 3 and 4. The use of air conditioning is notable in households 1 and 2, as seen by horizontal clusters of points.

### 2019-10-15 calendar-vis.tex
tweak figure 12
addressed the rest of the comments with Di

The purpose of the calendar-based visualization is to provide insights into people's daily schedules, relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia \citep{ped}. There are many sensors installed across the inner-city area, that count pedestrians every hour (Figure \ref{fig:ped-map}). Data from 43 sensors in 2016 is analyzed here. This data can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. Patterns relative to special events (such as public holidays and recurring cultural/sporting events) would be worth studying in comparison to regular days, but conventional displays of time series data may bury this detail.

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There are some existing works in this area. For example, \citet{VanWijkCluster1999} developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts weekdays and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. \citet{Jones-heatmap}, \citet{R-TimeProjection}, \citet{R-ggTimeSeries}, and \citet{R-ggcal} implemented some variants of calendar-based heatmaps in R packages: \textbf{TimeProjection}, \textbf{ggTimeSeries}, and \textbf{ggcal} respectively. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Color-encoding is also low on the hierarchy of optimal variable mapping \citep{cleveland1984graphical, lam2007overview}.

@@

The algorithm of transforming data for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data \citep{Wickham2012glyph}. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month conforms to a layout of 5 rows and 7 columns, where the top left is Sunday of week 1, and the bottom right is Monday of week 5 by default. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is deterministic given the year of the calendar. Months are of different lengths, ranging from 28 to 31 days. Some months could extend over six weeks, but for these months the last few days are wrapped up to the top row of the block for compactness, and because it is convention. The notation for creating these cells is as follows:

@@

{\centering \includegraphics[width=360pt,height=250pt]{img/month}

\caption{Illustration of the indexing layout for cells in a month, where \(k\) is day of the week, \(g\) is day of the month, \((i, j)\) indicates grid position.}\label{fig:month-diagram}
\end{figure}

@@

{\centering \includegraphics[width=360pt,height=250pt]{img/year-diagram}

\caption{Illustration of the indexing layout for months of one year, where \(M\) and \(N\) indicate number of rows and columns, \(b\) is a space parameter separating cells.}\label{fig:year-diagram}
\end{figure}

@@

\caption{The calendar plot of hourly foot traffic at Flagstaff Station, using line graphs. The disparities between weekday and weekend along with public holiday, are immediately apparent. The arrangement of the data into a \(3 \times 4\) monthly grid represents all the traffic in 2016. Note that the algorithm wraps the last few days in the sixth week to the top row of each month block for a compact layout, which occurs in May and October.}\label{fig:fs-2016}
\end{figure}

@@

Almost any type of plot can be shown in a calendar pane. Most of the full range of plotting capabilities in \textbf{ggplot2} is available. An example is shown in Figure \ref{fig:scatterplot}: the panes contain lag scatterplots, constructed with local scaling for each day at Flagstaff Station, where the lagged hourly count is assigned to the \texttt{x} argument and the current hourly count to the \texttt{y} argument. It indicates strong autocorrelation on weekends, and weak autocorrelation on work days. The V-shape in the weekday graphs arises when the next hour sees a substantial increase or decrease in counts. This is due to the morning and afternoon commuter patterns in Figure \ref{fig:fs-2016}: the adjacent hours are positively correlated when approaching to the peak hour but negatively correlated when moving away from the peak.

@@

{\centering \includegraphics[width=\textwidth]{figure/scatterplot-1}

@@

\caption{Lag scatterplots, with local scaling, in the daily calendar layout. Each hour's count is plotted against the previous hour's count at Flagstaff Station, to demonstrate the autocorrelation at lag 1. The correlation between them is more consistent on non-work days than work days.}\label{fig:scatterplot}
\end{figure}

@@

Linking calendar displays to other types of charts is valuable to visually explore the relationships between variables. An example can be found in the \textbf{wanderer4melb} shiny application \citep{R-wanderer4melb}. The calendar most naturally serves as a tool for date selection: by selecting and brushing the glyphs in the calendar, it subsequently highlights the elements of corresponding dates in other time-based plots. Conversely, selecting on weather data plots, linked to the calendar can help to assess if very hot/cold days and heavy rain, affect the number of people walking in downtown Melbourne. The linking between weather data and calendar displays is achieved using the common dates.

@@

\caption{Scatterplots of half-hourly energy usage by time of day, with hourly averages overlaid, faceted by household and type of day. All households have different weekdays versus weekends daily routines. On weekdays, household 1 wakes up early before 6, and household 2 around 6, followed by household 3 and 4. The use of air conditioning is notable in households 1 and 2, as seen by horizontal clusters of points.}\label{fig:hod}
\end{figure}

>>>>

The purpose of the calendar-based visualization is to provide insights into human activities, especially relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia \citep{ped}. There are many sensors installed across the inner-city area, that count pedestrians every hour (Figure \ref{fig:ped-map}). Data from 43 sensors in 2016 is analyzed here. This data can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. Patterns relative to special events (such as public holidays and recurring cultural/sporting events) would be worth studying in comparison to regular days, but conventional displays of time series data may bury this detail.

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There are some existing works in this area. For example, \citet{VanWijkCluster1999} developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts weekdays and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. Some variants of calendar-based heatmaps have been implemented in R packages: \textbf{TimeProjection} \citep{R-TimeProjection}, \textbf{ggTimeSeries} \citep{R-ggTimeSeries}, and \textbf{ggcal} \citep{R-ggcal}. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Color-encoding is also low on the hierarchy of optimal variable mapping \citep{cleveland1984graphical, lam2007overview}.

@@

The algorithm of transforming data for constructing a calendar plot uses modular arithmetic, similar to that used in the glyph map displays for spatio-temporal data \citep{Wickham2012glyph}. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month conforms to a layout of 5 rows and 7 columns, where rows and columns refer to weeks of the month and days of the week respectively. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is deterministic given the year of the calendar. Months are of different lengths, ranging from 28 to 31 days. Some months could extend over six weeks, but for these months the last few days are wrapped up to the top row of the block for compactness, and because it is convention. The notation for creating these cells is as follows:

@@

@@

@@

\caption{The calendar plot of hourly foot traffic at Flagstaff Station ranging from 0 to 6952, using line graphs. The disparities between weekday and weekend along with public holiday, are immediately apparent. The arrangement of the data into a \(3 \times 4\) monthly grid represents all the traffic in 2016. Note that the algorithm wraps the last few days in the sixth week to the top row of each month block for a compact layout, which occurs in May and October.}\label{fig:fs-2016}
\end{figure}

@@

Almost any type of plot can be shown in a calendar pane. Most of the full range of plotting capabilities in \textbf{ggplot2} is available. An example is shown in Figure \ref{fig:scatterplot}: the panes contain lag scatterplots, constructed with local scaling for each day at Flagstaff Station, where the lagged hourly count is assigned to the \texttt{x} argument and the current hourly count to the \texttt{y} argument. It indicates strong autocorrelation on weekends, and weak autocorrelation on work days. The V-shape in the weekday graphs arises when the next hour sees a substantial increase or decrease in counts.

@@

{\centering \includegraphics[width=0.75\linewidth]{figure/scatterplot-1}

@@

\caption{Lag scatterplots, with local scaling, in the daily calendar layout. Each hour's count is plotted against the previous hour's count at Flagstaff Station, to demonstrate the autocorrelation at lag 1. The correlation between them is more consistent on non-work days (blue) than work days (red).}\label{fig:scatterplot}
\end{figure}

@@

Linking calendar displays to other types of charts is valuable to visually explore the relationships between variables. An example can be found in the \textbf{wanderer4melb} shiny application \citep{R-wanderer4melb}. The calendar most naturally serves as a tool for date selection: by selecting and brushing the glyphs in the calendar, it subsequently highlights the elements of corresponding dates in other time-based plots. The linking between weather data and calendar displays is achieved using the common dates.

@@

\caption{Scatterplots of half-hourly energy usage by time of day, with hourly averages overlaid, faceted by household and type of day. On weekdays, household 1 wakes up early before 6, and household 2 around 6, followed by household 3 and 4. The use of air conditioning is notable in households 1 and 2, as seen by horizontal clusters of points.}\label{fig:hod}
\end{figure}
